---
title: "Practical 1"
author: "Matteo Larrode"
date: "2023-10-18"
output: html_document
---

# Week 1: Statistical Preliminaries

```{r setup}
library(tidyverse)

fishdata <- read_csv("data/fishdata.csv")
```

### a) Taking subsets and summarising variables

i) How many countries are predominantly Muslim?
```{r}
nrow(fishdata[fishdata$MUSLIM == 1, ])
```

ii) What percentage of countries are predominantly Muslim?
```{r}
nrow(fishdata[fishdata$MUSLIM == 1, ]) / nrow(fishdata) * 100
```

iii) How many countries have GDP in 1990 of above 3.0?
```{r}
nrow(fishdata[fishdata$GDP90LGN > 3,])
```

iv) How many countries are both Muslim and a former British colony?
```{r}
fishdata |> 
  filter(MUSLIM == 1 & BRITCOL == 1) |> 
  nrow()
```

v) How many countries have either average economic growth from 1975-98 of above 0.6% or GDP in 1990 of above 2.5?
```{r}
fishdata |> 
  filter(GRW7598P > .6 | GDP90LGN > 2.5) |> 
  nrow()
```

vi) Create a new dataset consisting only of countries that are both Muslim and a member of OPEC
```{r}
muslim_OPEC <- fishdata |> 
  filter(MUSLIM == 1 & OPEC == 1)
```

### b) What is the difference in mean Freedom House score between Muslim and Non-Muslim countries? 

Calculate it both by hand and using a regression, verifying that your answers are identical.

- By hand

```{r}
mean(fishdata$FHREVERS[fishdata$MUSLIM == 1]) - mean(fishdata$FHREVERS[fishdata$MUSLIM == 0])
```

- By regression
```{r}
summary(lm(FHREVERS ~ MUSLIM, fishdata))
```
Muslim countries score on average 2.20 points less democratic than non-Muslim countries.


### c) Is the difference in means in (b) likely to be biased? If so, in which direction and why?

This is only a bivariate relationship, without any controls. In reality Muslim countries may be different than non-Muslim countries in many other ways that also affect their level of democracy, e.g. their level of economic development. This suggests that the measure is likely to be biased.

Some obvious omitted variables, such as economic development, are likely to be positively correlated with a country’s democracy level but negatively correlated with being predominantly Muslim. This would imply that the difference in means is *biased downward*, because we have not accounted for the fact that predominantly Muslim countries are also poorer.

Other potential omitted variables are likely to be negatively correlated with the democracy level but positively correlated with being predominantly Muslim, including OPEC member- ship (and therefore being a major oil-producing country). For instance, the ‘resource curse’ theory in political science suggests that access to oil revenues allows governments to buy off citizens without introducing democracy. Again, this suggests downward bias from having failed to account for the fact that predominantly Muslim societies are also more likely to be OPEC members.


### T-tests

d) Conduct a t-test for the difference in means in (b) using R’s t.test() function. Is the difference statistically significant?

```{r}
t.test(fishdata$FHREVERS[fishdata$MUSLIM==1], fishdata$FHREVERS[fishdata$MUSLIM==0])
```
The difference is statistically significant at any significance level, because the t-statistic is -9.63 and the p-value is extremely close to zero.

e) Conduct the t-test again, this time coding it by hand. Confirm that your answer is identical to (d)

```{r}
d <- mean(fishdata$FHREVERS[fishdata$MUSLIM==1]) - mean(fishdata$FHREVERS[fishdata$MUSLIM==0])

se <- sqrt(
 var(fishdata$FHREVERS[fishdata$MUSLIM==1])/length(fishdata$FHREVERS[fishdata$MUSLIM==1]) +
   var(fishdata$FHREVERS[fishdata$MUSLIM==0])/length(fishdata$FHREVERS[fishdata$MUSLIM==0])
)

d/se

```

### f) Calculate: 

i) The percentage of Muslim countries that are former British colonies
```{r}
length(fishdata$MUSLIM[fishdata$BRITCOL==1 & fishdata$MUSLIM==1]) / sum(fishdata$MUSLIM)
```
_TIP_: here we use the sum() function to get the number of Muslim countries because the MUSLIM variable has been coded such that is the country is predominently Muslim, it gets a value of 1; therefore, non-predominently countries, with a value of 0 will not be counted in the sum().

ii) The percentage of non-Muslim countries that are former British colonies

Note that steps i) and ii) are akin to measuring the correlation between being a Muslim country and being a former British colony. We do it this way because a correlation coefficient is defined only for two continous variables, and these are both binary.
```{r}
length(fishdata$MUSLIM[fishdata$BRITCOL==1 & fishdata$MUSLIM==0]) / (length(fishdata$MUSLIM)-sum(fishdata$MUSLIM))
```

iii) The correlation between being a former British colony and Freedom House score, controlling for being Muslim.
```{r}
summary(lm(FHREVERS ~ MUSLIM + BRITCOL, data=fishdata))
```
- 15.9% of Muslim countries are former British colonies;
- 23.1% of non-Muslim countries are former British colonies
-> Being a Muslim country and being a former British colony are negatively correlated. 

However, the correlation between being a former British colony and the Freedom House score, controlling for being Muslim, is virtually zero. The coefficient on BRITCOL is very small (bearing in mind that the freedom house score ranges from 1 to 7) and not statistically significant.

As a result, the coefficient on MUSLIM is virtually unchanged compared to the difference in means we estimated in (d) and (e). Being a British colony is not an omitted variable here.

